\mytitle

### 2023-05-31 paper-seb-main.tex
Revisions getting closer to sending to Computer Journal, and renaming the repo

\input paper-seb-macros.tex
\input generated/info-for-main.tex
\input generated/page-lengths.tex

@@

\long\def\ignore#1{}

@@

%\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
%&\texttt{harold@thimbleby.net}\\
%&ORCID 0000-0003-2222-4243
%\end{tabular}

@@

\begin{abstract}
\justifying
%\color{red}
%This is a placeholder for the revisions that are almost complete. I have uploaded it now in case the computer system would lose all record of the submission --- I have emailed to ask for a short extension, but with the offices shut, there is no chance of any negotiation. Thanks for your understanding.
%\color{black}\vskip .2cm

\noindent
Scientists rarely rigorously ensure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny. Even if available, scientists rarely provide adequate documentation to understand or use their code reliably. When code used in science avoids adequate scrutiny, published science will be unreliable and unverifiable because results --- data, graphs, images, inferences, and so forth --- informed or generated by the code will themselves be unreliable and unverifiable.

@@

%\emph{Problems:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit, even when code is made available. Both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects. Code is fallible, so any interpretation that relies on code is also fallible. When the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized. In turn, scientific claims cannot be properly scrutinized.

%\emph{Methodology:} From the perspective of Software Engineering, this paper critiques the quality (and accessibility) of coding in computational science, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions. 
%The paper reports a pilot survey of peer-reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.

%Code can be improved using Software Engineering. This paper argues for specific solutions:

%\emph{Results:} %Journals do not always have a code policy, and many have a relaxed approach to code disclosure.

%\noindent\emph{Solutions:}

@@

%\begin{quote}
\def\seprule{\vskip 2ex\hrule\vskip 2ex}
\seprule
\raggedright
\def\zq{\setbox0=\hbox{``}\hskip -0\wd0\copy0}
\zq Science is what we understand well enough to explain to a computer.''\\
\hfill {\rm Donald E. Knuth in $A=B$ \cite{a=b}}\\

@@

\seprule
%\end{quote}
\end{abstract}
\maketitle

@@

This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research because of its significance for informing and driving public health responses to the Covid pandemic. A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey. In the survey, \emph{no\/} papers claimed or provided evidence that their code was adequately tested or rigorously developed; \emph{none\/} used methodologies like RAP or \RAPstar\ (described below). Only one paper mentioned any Software Engineering methods, albeit simplistic and without technical details.

@@

Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having that consensus in turn depended on reproducible thermometer measurements; for example, scientists working in different places needed to know they were working with the ``same temperatures'' yet there was for a long period no consensus on what that meant. Contributing to reliable science depended on a thorough understanding of principles, including gradually fixing the confounding factors that were misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.} 
Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, that follow international quality standards.

@@

In the ``old days'' such numbers (\the\dataN; \the\countAuthors; etc) quoted above would have been manually worked out, eyeballed, typed up, and then (hopefully) double-checked by the authors. Instead, in this paper, those numbers (and many more) were computed automatically. They were inserted into the text of this paper automatically, and are auditable back to their sources. Furthermore, they will update automatically --- with no further work from the author --- when the data or calculations change.

@@

where the \the\countAuthors\ is the calculated number, save that line of text to a file, which is then import it into the paper where it is typeset in the normal way. Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the actual value is at the time. Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done more realistically elsewhere in this paper (for instance, see footnote \ref{name-example-footnote}).

The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results in the paper also update automatically, and all the statistics, graphs, and analysis reported in the paper remain correct \emph{with no further work from the authors}. Indeed, if an author corrects a mistake, the correction will apply automatically to all future revisions.

@@

A taut review of such systems is Perkel \cite{perkel-review}, whereas \cite{maneage} discusses in-depth the design trade-offs of one powerful approach, Maneage; the paper \cite{maneage} includes (in an appendix) a substantial and useful literature review. Here four representative tools (WEB, \emph{Mathematica}, Jupyter, and knitr) serve to sample the variety of approaches that are available:

@@

The authors now have a maintenance problem that PRISMA does not address: where are the numbers that have changed, and what should they be changed to? Doing a search-and-replace, whether automated or by hand, is fraught with difficulties. What happens if $2000$ is used for some other purposes as well? What happens if some of the $2000$ values are written as $2,000$ or as $2\hskip 0.3em 000.0$? What happens if some $2000$ are year dates and are changed incorrectly? Then there are the Human Factors: slips and errors will happen in this workflow anyway \cite{engima}. Typos, slips during cut-and-paste, and other errors are common. Similar iterative revision cycles happen with any paper, not just with systematic reviews. PRISMA ignores the methodological problems of using code (such as the example earlier in this paragraph).

The irony is that PRISMA says nothing about how to ensure the results of a survey are correctly and reliably presented, despite this being one of PRISMA's explicit motivations. PRISMA explicitly warns about methodological issues, but ignores that poorly managed code raises methodological issues that will undermine the validity of research it supports. PRISMA reinforces the culture that code is trivial and incidental to research.

@@

\subsection{Case study: Pandemic modeling}
\label{section-pandemic-modeling}
A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modeling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer-reviewers, there is no role for code or computation as such. It appears that the new insights, advances, rigor, and problems that computers bring to research were not considered relevant.

A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}. The review \cite{diagnosis-reviews} itself ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted regardless of whether TRIPOD guidelines are followed. Indeed, TRIPOD guidelines ignore code completely.

It should be noted that flowcharts, which the review \cite{diagnosis-reviews} did consider, are programs intended for human use. Flowcharts, too, should be designed as carefully as code, for exactly the same reasons.

A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} originally developed for modeling H5N1 in Thailand, when it did not model air travel or other factors required for later western COVID-19 modeling. The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.

A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,

@@

This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, and a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times. The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022. (Typical $h$ values vary by discipline.)} of 93. Ferguson should be well aware of the standards of coding used in his own field. Arguably, then, this comment quoted above is representative of the culture of the field.

Ferguson's admission is tantamount to saying that the published scientific findings are and need not be reproducible.\footnote{A constructive discussion of Software Engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.}

Lack of reproducibility is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with its different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.

@@

If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results. In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.

@@

Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compliers, libraries, or hardware. In fact, in any area where reliable programming is required in a C-like language, a special dialect such as MISRA C is preferred: MISRA C manages the serious design flaws of C that otherwise make it too unreliable \cite{misra}. Alternatively, a high integrity programming language, unrelated to C, such as SPARK Ada \cite{spark}, or modern languages (many related to the ``ML family'') like OCaml, F*, Haskell \cite{haskell} could be used. These languages have steeper learning curves; however their key benefit is that correct programs are far more likely and are \emph{much\/} faster to write. (The \supplement\ discusses these issues further.)

@@

\begin{quote}\raggedright
{\sf\setbox0=\hbox{``}
\hskip -\wd0\box0 
However, none of the criticisms of the code affects the mathematics or science of the simulation''} \cite{thumbs-up}
\end{quote}

@@

The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirmed that the code could be run again and produced comparable results (compared using an Excel spreadsheet). This weak test would pass provided the runs gave the same invalid answers --- it is not usefully a stronger test than just checking that the code compiles.

Running code to obtain results claimed in a paper is a very weak test, and anyway one that should have been checked automatically during paper preparation and submission. If you do not know what you are reproducing, as is the case in this \emph{Nature\/} paper, there is little scientific value in doing so.

Unfortunately, the terms reproducibility, replicability, and repeatability, have similar meanings in English but have been used in different specific technical ways by different authors. In \cite{codecheck,thumbs-up} the reproduction amounted to just re-running the original code. It is certainly essential to establish that a paper's code can be run, as non-working code cannot support any claims in a paper; if the original code runs this confirms a basic level of access for the wider scientific community. But a more realistic criterion than basic reproduction in this sense is whether an \emph{independently\/} developed model developed from the same paper(s) specifications produces equivalent results (called $N$-version programming, a standard Software Engineering practice \cite{NVP}) like public health surely requires as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues. However, much stronger scrutiny of code than ``reproduction'' is required to answer essential questions including:

\newcount \enum
\enum=1
\renewcommand{\theenumi}{\ifnum \enum<10 \hphantom{0}\fi
\the\enum
\global\advance \enum by 1}
\begin{enumerate}\raggedright
\item Is the code valid: does it do what the paper claims?\footnote{Of course, the underlying science may be wrong to, so it is useful to distinguish \emph{internal validity\/} and \emph{external validity}. Internal validity occurs if the code does what the paper claims; external validity occurs if the code represents correct science which the paper may have interpreted incorrectly.}
\item Do other scientists, including reviewers and the authors, understand the code?
\item Does the code implement the methods described in the paper?
\item Has the code been over-fitted or tweaked to support specific claims in the paper?'
\item Is there a definitive version of code?
\item Is the code controlled and signed?
\item What limitations does the code have?
\item Was the code developed to any standard, and does it comply to that standard?
\item How does the code protect against data, coding, and human error?
\item Was the code tested adequately?
\item Does the code depend on arbitrary parameters, data, or code to over-fit to obtain the published results?
\item Is the code documented adequately, so we know what it is trying to do, and how?
\item \ldots\ and so forth --- questions are numbered for reference. 
\end{enumerate}

@@

It is well-known that documenting code helps impro
Some papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming. Pseudo-code looks deceptively like real code that might be copied to try to reproduce it, but pseudo-code introduces invisible and unknown simplifications. Pseudo-code, properly used, can give a helpful impression of the overall approach of an algorithm, certainly, but pseudo-code alone is not a surrogate for code: using it instead of making actual code available is worse than not publishing code at all (see \cite{chinese}). Pseudo-code is too vague to help anyone scrutinize a model; moreover, pseudo-code may mask over-fitting in code used that is not explicit in the pseudo-code.

@@

Epidemiology has a high profile because of the COVID pandemic, but the problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication. But other examples that were not rushed include a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model}, which provides no details of its code. Its supplementary material, which might have provided code, no longer exists.

There are many other areas of computational science that are equally if not more critical, and many will have longer-lasting impact. Climate change modeling is one such example that will have an impact long beyond the COVID pandemic.

@@

There is a trade-off to balance when and how much effort to put into reproducibility. The trade-off is comparable to trade-offs in using statistics --- the author may realize at a later stage that a ``significant'' result depends on designing the experiment appropriately for the intended statistical claims, but now making the claim rigorously requires revising the methodology and probably improving the type of analysis too. What was an easier route to take earlier now causes a challenging and costly revision. However, since mature scientists recognize the importance of correct statistics, statistics is positioned at the forefront of their work rather than delayed and sorted out at greater cost later. Statistics is routinely in the undergraduate syllabus, so most scientists consciously make appropriate trade-offs minimizing statistical technical debt.

@@

\subsection{Approaches to further work}
Encouraging and informing the improvement of science and, specifically the reproducibility of science that relies on code, were the main aims of this paper. Further work should research the efficiency, effectiveness, and quality of the various ideas proposed, such as \RAPstar\ and Software Engineering Boards, and propose and evaluate more ideas.

@@

The present paper did not assess the correctness of code used in papers for several reasons. Code was not documented well-enough (even with high-level discussion in the papers) to know what ``correct'' would mean, and no papers performed adequate tests, let alone provided adequate test material for independent verification (see Table \ref{table-summary}); moreover, installing the software environments to build and test the systems --- different environments for each paper --- was too onerous, even when those environments were specified. The implication is that people within the relevant fields, especially referees, should promote standardized software environments to help increase the rate of reproduction and verification of results. As a matter for further research, then, it is important to develop, assess, and promote effective shared online (e.g., cloud) environments, perhaps with discipline-specific solutions, so that development and test environments are standardized, powerful enough, and sufficiently accessible.

\section{Conclusions}
A pandemic creates unprecedented pressure and exposes problems in scientific methodology. During the COVID crisis, code led epidemiological modeling, implemented track and trace and caused problems \cite{excel-fiasco}, modeled mutation pressures against vaccine shortages \cite{science-delays}, and more. Code drove public policy. Code had a direct impact on the quality of life.

While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview} about his high-profile COVID pandemic modeling, the evidence reviewed here suggests that scientific coding practice is inadequate in every field, but particularly worrying in the context of the extreme pressures of managing a pandemic in real time. As science becomes more and more reliant on computers, we need to correspondingly improve the quality of code, the quality of code policies, the quality of Software Engineering, and the quality of all scientists' understanding of computation and how to manage its unlimited complexity.

@@

Software Engineering Boards (SEBs), as proposed in this paper, are an initial, straightforward, constructive, and practical way to support and improve computer-based science. If nothing else, the idea of SEBs is something to criticize and improve.

@@

\section*{Supporting information}
\newcount\csrefcount \csrefcount=0
\def\csref{\global\advance\csrefcount by 1}
\long\def\ethics#1{\paragraph*{Ethics}#1}
\long\def\ack#1{\paragraph*{Acknowledgments}#1}
\long\def\dataaccess#1{\paragraph*{Data and code access}#1}
\long\def\aucontribute#1{\paragraph*{Author contribution}#1}
\long\def\competing#1{\paragraph*{Competing interests}#1}
\long\def\funding#1{\paragraph*{Funding}#1}

@@

%\immediate\write\@auxout{\string \newcount \string \savedlastReferenceNumber
%\string \savedlastReferenceNumber=\the\savedlastReferenceNumber}

>>>>

\input paper-seb-macros.tex
\input generated/page-lengths.tex

@@

@@

@@

\begin{abstract}
\justifying
\noindent
Scientists rarely rigorously ensure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny. Even if available, scientists rarely provide adequate documentation to understand or use their code reliably. When code used in science avoids adequate scrutiny, published science will be unreliable and unverifiable because results --- data, graphs, images, inferences, and so forth --- informed or generated by the code will themselves be unreliable and unverifiable.

@@

@@

\def\seprule{\vskip 2ex\hrule\vskip 2ex}
\seprule
\raggedright
\def\zq{\setbox0=\hbox{``}\hskip -0\wd0\copy0}
\zq Science is what we understand well enough to explain to a computer.''\\
\hfill {\rm Donald E. Knuth in $A=B$ \cite{a=b}}\\

@@

\seprule
\end{abstract}
\maketitle

@@

This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research because of its significance for informing and driving public health responses to the COVID-19 pandemic. A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey. In the survey, \emph{no\/} papers claimed or provided evidence that their code was adequately tested or rigorously developed; \emph{none\/} used methodologies like RAP or \RAPstar\ (described below). Only one paper mentioned any Software Engineering methods, albeit simplistic and without technical details.

@@

Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having that consensus in turn depended on reproducible thermometer measurements; for example, scientists working in different places needed to know they were working with the ``same temperatures'' yet there was for a long period no consensus on what that meant. Contributing to reliable science depended on a thorough understanding of principles, including gradually fixing the confounding factors that were misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the volume of mercury also increases with temperature (but at a different rate), so the volume measurement is inaccurate.} 
Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, that follow international quality standards.

@@

In the ``old days'' such numbers (\the\dataN; \the\countAuthors; etc) quoted above would have been manually worked out, eyeballed, typed up, and then (hopefully) double-checked by the authors. Instead, in this paper, those numbers (and many others) were computed automatically, and were then inserted into the text of this paper automatically. They are auditable back to their sources. Furthermore, they will update automatically --- with no further work from the author --- when the data or calculations change.

@@

where the \the\countAuthors\ is some number calculated from the relevant data.

Such a line of text is then saved to a file, which is then imported into the paper so the value is named and can be used easily and reliably. Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the actual value is at the time --- it will update automatically whenever the data or analysis is improved. Some easy \LaTeX\ coding can then present the numbers in the author's or publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done more realistically elsewhere in this paper (for instance, see footnote \ref{name-example-footnote}).

The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results written up in the paper \emph{also\/} update automatically, and all the statistics, graphs, and analysis reported in the paper update and remain correct \emph{with no further work from the authors}. Indeed, if an author corrects a mistake, the correction will apply automatically to all future revisions.

@@

A taut review of such systems is Perkel \cite{perkel-review}, whereas \cite{maneage} discusses in-depth the design trade-offs of one powerful approach, Maneage; the paper \cite{maneage} includes  a substantial and useful literature review (in its appendix). Here, four representative tools (WEB, \emph{Mathematica}, Jupyter, and knitr) serve to sample the variety of approaches that are available:

@@

The authors now have a maintenance problem that PRISMA does not address: where are the numbers that have changed, and what should they be changed to? Doing a search-and-replace, whether automated or by hand, is fraught with difficulties. What happens if $2000$ is used for some other purposes as well? What happens if some of the $2000$ values are written as $2,000$ or as $2\hskip 0.3em 000.0$, or if a number containing the digits 2000, like 12000, is changed? What happens if some $2000$ are year dates and are changed incorrectly?

Then there are the Human Factors: slips and errors will happen in this workflow anyway \cite{engima}. Typos, slips during cut-and-paste, and other errors are common. Similar iterative revision cycles happen with any paper, not just with systematic reviews.

PRISMA, like many such standards, ignores the methodological problems of using code such as the issues raised above.

The irony, then, is that PRISMA says nothing about how to ensure the results of a survey are correctly and reliably presented, despite this being one of PRISMA's explicit motivations. PRISMA explicitly warns about methodological issues, but ignores that poorly managed code raises methodological issues that will undermine the validity of research it supports. PRISMA reinforces the culture that code is trivial and incidental to research.

@@

\subsection{Case study: Pandemic modeling}
\label{section-pandemic-modeling}
For an excellent review of the extreme pressures under which scientists were working during the COVID-19 pandemic (but nothing about the role of computers!), see \cite{preventable-book} --- which, while referring to some failed mRNA vaccines, makes an important point on recovering from ``failed'' science:

\begin{quote}\raggedright
{\sf\setbox0=\hbox{``}
\hskip -\wd0\box0
Such is the beauty of science: even failed attempts are a step towards more information and progress forward.''} \cite{preventable-book}
\end{quote}

But progress, if any, would be chaotic if it was not possible to scrutinize exactly what science has failed. It undermines progress if science and the code it relies on are not both accessible and adequately defined. It is as misleading as getting the statistics or mathematics wrong.

By focusing on influential science and coding undertaken to inform public health policies during a pandemic emergency, this section now focuses on an area where reliable and high quality science using code open to interdisciplinary scrutiny was very obviously required. Note that pandemic model correctness is a secondary concern here; correctness is not addressed as without informed scrutiny correctness cannot even be assessed.

A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modeling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer-reviewers, there is no role for code or computation as such. It appears that the new insights, advances, rigor, and problems that computers bring to research were not considered relevant.

A systematic review \cite{diagnosis-reviews} of published COVID-19 models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}. The review \cite{diagnosis-reviews} itself ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted regardless of whether TRIPOD guidelines are followed. Indeed, TRIPOD guidelines ignore code completely.

It should be noted that flowcharts, which the review \cite{diagnosis-reviews} did consider, are graphical programs intended for human use. Flowcharts, too, should be designed as carefully as code, for exactly the same reasons.

A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel}, which influenced UK COVID-19 public health strategies, uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} originally developed for modeling H5N1 in Thailand, when it did not model air travel or other factors required for later western COVID-19 modeling. The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.

A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented code. As Ferguson, the original code author, noted in an interview,

@@

The admission above is tantamount to saying that the published scientific findings are and need not be reproducible.\footnote{A constructive discussion of Software Engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.}

The comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications involving computational modeling, with a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times. The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022. (Typical $h$ values vary by discipline.)} of 93, and ``one of the top scientists advising the government on its response to the coronavirus crisis'' working in the UK's Scientific Advisory Group for Emergencies (SAGE) group \cite{ferguson-resign}.

Ferguson's code must be representative of best practice at a time when the stakes were high and reliability was known to be essential; and if not representative of best practice, at least representative of accepted practice both in Ferguson's team, the field of epidemiology more widely, as well as with members of the high-powered interdisciplinary SAGE group. It is therefore instructive to explore the larger story around this science that uses code.

Lack of reproducibility is problematic, especially as the model code would have required many non-trivial modifications to update it for COVID-19 with its different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.

@@

If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results (i.e., over-fitting, see section \ref{over-fit}); worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results. In contrast, if code is properly documented, the documentation defines the original intentions (including, where appropriate, formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.

@@

Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compilers, libraries, or hardware. In fact, in any area where reliable programming is required in a C-like language, a special dialect such as MISRA C is preferred: MISRA C manages the serious design flaws of C that otherwise make it too unreliable \cite{misra}. Alternatively, a high integrity programming language, unrelated to C, such as SPARK Ada \cite{spark}, or modern languages (many related to the ``ML family'') like OCaml, F*, Haskell \cite{haskell} could be used. These languages have steeper learning curves; however their key benefit is that correct programs are far more likely and are \emph{much\/} faster to write. (The \supplement\ discusses these issues further.)

@@

\begin{quote}\raggedright
{\sf\setbox0=\hbox{``}
\hskip -\wd0\box0 
However, none of the criticisms of the code affects the mathematics or science of the simulation.''} \cite{thumbs-up}
\end{quote}

@@

The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirms that the code can be run again and produce comparable results. As Eglen says,

\begin{quote}\raggedright
{\sf\setbox0=\hbox{``}
\hskip -\wd0\box0 Each run generated a tab-delimited file in the output folder. Two R scripts provided by Prof Ferguson were used to summarise these runs into two summary files [\ldots] These files were compared against the values generated by Prof Ferguson [\ldots] The results were found to be identical. Inserting my results into his Excel spreadsheet generated the same pivot tables. The codecheck found that: `Small variations (mostly under 5\%) in the numbers were observed [\ldots]'\,''} \cite{codecheck}
\end{quote}

This test would pass provided the runs gave the same answers regardless of whether the answers are correct --- it is not a usefully stronger test than just checking that the code compiles. The comparison relied on running (apparently) unchecked R code to summarize the data, which is potentially misleading unless the published results \cite{ICmodel} exclusively relied on the same summary code. In general, reproducing code results, even done formally, does not scrutinize the science, as  \cite{codecheck-certificate} makes clear.

Running code just to obtain results claimed in a paper is a weak test, and anyway one that should be checked routinely during paper preparation and submission. However, in this case the reproduction involved a community effort that also refactored and improved the code, which added value to the code and usefully improved its generality \cite{codecheck}. The reproduction effort was also certified \cite{codecheck-certificate}, which is the sort of evidence of quality assurance processes that arguably should be required before publication, particularly for critical code such as public health modeling. The publicity of this story and the certificate will certainly raise the profile of the scientific value of independent review of code.

Unfortunately, the terms reproducibility, replicability, and repeatability, have similar meanings in English but have been used in different specific technical ways by different authors. In \cite{codecheck,thumbs-up} the reproduction amounted to just re-running the original code. It is certainly essential to establish that a paper's code can be run, as non-working code cannot support any claims in a paper; if the original code runs this confirms a basic level of access for the wider scientific community, and this can be formally certified \cite{codecheck-certificate} so it is appreciated by the community. But a more realistic criterion than basic reproduction in this sense is whether an \emph{independently\/} developed model developed from the same paper(s) specifications produces equivalent results (called $N$-version programming, a standard Software Engineering practice \cite{NVP}) like public health surely requires as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues.

In general, much stronger scrutiny of code than ``reproduction'' is required to answer essential questions (numbered below for reference) including:

\newcount \enum
\enum=1
\renewcommand{\theenumi}{\ifnum \enum<10 \hphantom{0}\fi
\the\enum
\global\advance \enum by 1}
\begin{enumerate}\raggedright
\item Is the code valid: does it do what the paper claims?\footnote{Of course, the underlying science may be wrong to, so it is useful to distinguish \emph{internal validity\/} and \emph{external validity}. Internal validity occurs if the code does what the paper claims; external validity occurs if the code represents correct science which the paper may have interpreted incorrectly.}
\item Do other scientists, including reviewers and the authors, understand the code?
\item Does the code implement the methods described in the paper?
\item Has the code been over-fitted or tweaked to support specific claims in the paper?
\item Is there a definitive version of code?
\item Is the code controlled and signed?
\item What limitations does the code have?
\item Was the code developed to any standard, and does it comply to that standard?
\item How does the code protect against data, coding, and human error?
\item Was the code tested adequately?
\item Does the code depend on arbitrary parameters, data, or code to over-fit to obtain the published results?
\item Is the code documented adequately, so we know what it is trying to do, and how?
\item \ldots\ and so forth. 
\end{enumerate}

@@

It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost. The revised code has now been published, and it too has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.

Some papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming. Pseudo-code looks deceptively like real code that might be copied to try to reproduce it, but pseudo-code introduces invisible and unknown simplifications. Pseudo-code, properly used, can give a helpful impression of the overall approach of an algorithm, certainly, but pseudo-code alone is not a surrogate for code: using it instead of making actual code available is worse than not publishing code at all (see \cite{chinese}). Pseudo-code is too vague to help anyone scrutinize a model; moreover, pseudo-code may mask over-fitting in code used that is not explicit in the pseudo-code.

@@

Epidemiology has a high profile because of the COVID-19 pandemic, but the problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication. But other examples that were not rushed include a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model}, which provides no details of its code. Its supplementary material, which might have provided code, no longer exists.

There are many other areas of computational science that are equally if not more critical, and many will have longer-lasting impact. Climate change modeling is one such example that will have an impact long beyond the COVID-19 pandemic.

@@

There is a trade-off to balance when and how much effort to put into reproducibility. The trade-off is comparable to trade-offs in using statistics --- the author may realize at a later stage that a ``significant'' result depends on designing the experiment appropriately for the intended statistical claims, but now making the claim rigorously requires revising the methodology and probably improving the type of analysis too. What was an easier route to take earlier now causes a challenging and costly revision. However, since mature scientists recognize the importance of correct statistics, statistics is positioned at the forefront of their work rather than delayed and sorted out at greater cost later. Because of its recognized role, statistics is routinely in the undergraduate syllabus, so most scientists consciously make appropriate trade-offs minimizing statistical technical debt.

@@

\subsection{Approaches to further work}
Encouraging and informing the improvement of science and, specifically the reproducibility of science that relies on code, were the main aims of this paper. This paper raised problems and suggested some possible solutions: there are solutions, and better ones may yet be found. Although further work is desirable, any contributions can help improve science; not everything needs doing before we start.

Further work should research the efficiency, effectiveness, and quality of the various ideas proposed, such as \RAPstar\ and Software Engineering Boards, and propose and evaluate more ideas.

@@

The present paper did not assess the correctness of code used in papers (explicitly so in section \ref{section-pandemic-modeling}) for several reasons. Code was not documented well-enough (even with high-level discussion in the papers) to know what ``correct'' would mean, and no papers performed adequate tests, let alone provided adequate test material for independent verification (see Table \ref{table-summary}); moreover, installing the software environments to build and test the systems --- different environments for each paper --- was too onerous, even when those environments were specified. The implication is that people within the relevant fields, especially referees, should promote standardized software environments to help increase the rate of reproduction and verification of results. As a matter for further research, then, it is important to develop, assess, and promote effective shared online (e.g., cloud) environments, perhaps with discipline-specific solutions, so that development and test environments are standardized, powerful enough, and sufficiently accessible.

\section{Conclusions}
A pandemic creates unprecedented pressure and exposes problems in scientific methodology. During the COVID-19 crisis, code led epidemiological modeling, implemented track and trace and caused problems \cite{excel-fiasco}, modeled mutation pressures against vaccine shortages \cite{science-delays}, and more. Code drove public policy. Code had a direct impact on the quality of life.

While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview} about his high-profile COVID-19 pandemic modeling, the evidence reviewed here suggests that scientific coding practice is inadequate in every field, but particularly worrying in the context of the extreme pressures of managing a pandemic in real time. As science becomes more and more reliant on computers, we need to correspondingly improve the quality of code, the quality of code policies, the quality of Software Engineering, and the quality of all scientists' understanding of computation and how to manage its unlimited complexity.

@@

Software Engineering Boards (SEBs), as proposed in this paper, are an initial, straightforward, constructive, and practical way to support and improve code- and computer-based science. If nothing else, the idea of SEBs is something to criticize and improve.

@@

\section*{Supporting information}
\newcount\csrefcount \csrefcount=0
\def\csref{\global\advance\csrefcount by 1}
\long\def\ethics#1{\paragraph*{Ethics}#1}
\long\def\ack#1{\paragraph*{Acknowledgments}#1}

\newdimen\saveparindent \saveparindent=\parindent
\newdimen\highlightwidth \highlightwidth=\columnwidth
\advance \highlightwidth by -.5em

\long\def\highlight#1{\vskip 1em\noindent\colorbox{yellow}{\noindent\begin{minipage}{\highlightwidth}
\parindent=\saveparindent
#1\end{minipage}}}
\long\def\dataaccess#1{\highlight{\paragraph*{Data and code access}#1}}

\long\def\aucontribute#1{\paragraph*{Author contribution}#1}
\long\def\competing#1{\paragraph*{Competing interests}#1}
\long\def\funding#1{\paragraph*{Funding}#1}

@@
